Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-03-20 paper.qmd
Updated Figure 9

trimesh_two_curvy_umap <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000",
               alpha = 0.5,
               size = 0.5) +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  xlim(c(0, 1)) +
  ylim(c(0, 1)) +
  interior_annotation("a1", sc_ltr_pos, cex = 2)

@@

@@

benchmark_two_curvy <- 7 * a1

@@

benchmark_two_curvy <- 4 * bin_width

@@

text_df <- tibble::tibble(
  x = c(0.035, 0.15),
  y = c(10, 10),
  text = c("c", "b")
)

distance_points_umap <- ggplot(
  distance_two_curvy_dist,
  aes(x = distance,
      y = dist_highd)) +
  geom_point(alpha = 0.5) +
  geom_text(
    data=text_df,
    aes(x=x, y=y,
        label=text),
    colour="#bdbdbd",
    size = 10) +
  geom_vline(xintercept = 6.5 * bin_width,
             linetype="solid",
             color = "#bdbdbd",
             linewidth=1) +
  geom_vline(xintercept = 2 * bin_width,
             linetype="solid",
             color = "#bdbdbd",
             linewidth=1) +
  ylab(expression(d^{(4)})) +
  xlab(expression(d^{(2)})) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        panel.grid.major.x = element_blank(),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
```

@@

proj_obj1 <- get_projection(projection = projection, 
                            proj_scale = 1, 
                            scaled_data = scaled_two_curvy_data, 
                            scaled_data_model = scaled_two_curvy_data_model, 
                            distance_df_small_edges = distance_df_small_edges_two_curvy, 
                            axis_param = list(limits = 0.6, 
                                              axis_scaled = 2,
                                              axis_pos_x = -0.5, 
                                              axis_pos_y = -0.5, 
                                              threshold = 0.04))

@@

```{r}
#| label: fig-rm-lg
#| fig-cap: "The choice of long edge removal significantly impacts the model's accuracy. \\gD{} Vs $7\\text{-}D$ distances between the bin centroids, where the model is fitted with UMAP for the non-linear clusters data. The two benchmarks for long edge removal: (b) benchmark = 6.5$a_1$ and (c) benchmark = 2$a_1$, where $a_1$ = 0.03 are made. The wireframes in \\gD{} and a projection from $7\\text{-}D$ are shown with Delaunay triangulation (a1, a2) and the two benchmark values (b1, b2, c1, c2). With the benchmark set at 6.5$a_1$, long edges exist within the C-shaped cluster, which distrupts the model. In contrast, using 2$a_1$ is a better option, as it effectively captures the non-linear shapes of the clusters. To generate an accurate model that captures the data structure, the choice of long edge removal does matter."
#| fig-height: 15
#| fig-width: 15

>>>>

trimesh_two_curvy_umap <- ggplot() +
  geom_segment(data = tr_from_to_df_two_curvy,
               aes(
                 x = x_from,
                 y = y_from,
                 xend = x_to,
                 yend = y_to),
               colour = "#000000",
               alpha = 0.5,
               linewidth = 0.5) +
  geom_point(data = umap_two_curvy_scaled,
             aes(
               x = UMAP1,
               y = UMAP2
             ),
             color = "#636363",
             alpha = 0.5,
             size = 0.5
  ) +
  coord_equal() +
  xlim(c(0, 1)) +
  ylim(c(0, 1)) +
  interior_annotation("a1", sc_ltr_pos, cex = 2)

@@

# With the computed benchmark value
# distance_df_small_edges_two_curvy <- distance_two_curvy |>
#   filter(distance < benchmark_two_curvy)

@@

benchmark_two_curvy <- 8 * a1

@@

benchmark_two_curvy <- 2.5 * bin_width

@@

text_df <- tibble::tibble(
  x = c(0.12, 0.43),
  y = c(10, 10),
  text = c("c", "b")
)

distance_points_umap <- ggplot(
  distance_two_curvy_dist,
  aes(x = distance,
      y = dist_highd)) +
  geom_point(alpha = 0.5) +
  geom_text(
    data=text_df,
    aes(x=x, y=y,
        label=text),
    colour="#bdbdbd",
    size = 10) +
  geom_vline(xintercept = 8 * bin_width,
             linetype="solid",
             color = "#bdbdbd",
             linewidth=1) +
  geom_vline(xintercept = 2.5 * bin_width,
             linetype="solid",
             color = "#bdbdbd",
             linewidth=1) +
  ylab(expression(d^{(4)})) +
  xlab(expression(d^{(2)})) +
  theme_minimal() +
  theme(aspect.ratio = 0.75,
        panel.border = element_rect(fill = 'transparent'),
        plot.title = element_text(size = 20, hjust = 0.5, vjust = -0.5),
        panel.grid.major.x = element_blank(),
        axis.ticks.x = element_line(),
        axis.ticks.y = element_line(),
        axis.title = element_text(size = 15),
        axis.text = element_text(size = 12))
```

@@

proj_obj1 <- get_projection(projection = projection, 
                            proj_scale = 1, 
                            scaled_data = scaled_two_curvy_data, 
                            scaled_data_model = scaled_two_curvy_data_model, 
                            distance_df_small_edges = distance_df_small_edges_two_curvy2, 
                            axis_param = list(limits = 0.6, 
                                              axis_scaled = 2,
                                              axis_pos_x = -0.5, 
                                              axis_pos_y = -0.5, 
                                              threshold = 0.04))

@@

```{r}
#| label: fig-rm-lg
#| fig-cap: "The choice of long edge removal significantly impacts the model's accuracy. \\gD{} Vs $7\\text{-}D$ distances between the bin centroids, where the model is fitted with UMAP for the non-linear clusters data. The two benchmarks for long edge removal: (b) benchmark = 8$a_1$ and (c) benchmark = 2.5$a_1$, where $a_1$ = 0.03 are made. The wireframes in \\gD{} and a projection from $7\\text{-}D$ are shown with Delaunay triangulation (a1, a2) and the two benchmark values (b1, b2, c1, c2). With the benchmark set at 6.5$a_1$, long edges exist within the C-shaped cluster, which distrupts the model. In contrast, using 2$a_1$ is a better option, as it effectively captures the non-linear shapes of the clusters. To generate an accurate model that captures the data structure, the choice of long edge removal does matter."
#| fig-height: 15
#| fig-width: 15
